\documentclass[11pt,a4paper]{article}

\usepackage[T1]{fontenc}
\usepackage{isabelle,isabellesym}
\usepackage{amsmath,amssymb}
\usepackage{hyperref}

\begin{document}

\title{First-Order Methods for Smooth Convex Optimization in Isabelle/HOL}
\author{Feier Lyu}
\maketitle

\begin{abstract}
This entry develops reusable Isabelle/HOL infrastructure for first-order
methods in smooth convex optimization.  The central application is
projected-gradient descent, but the development is organized around
general-purpose interfaces for gradients, first-order convexity certificates,
smooth quadratic upper bounds, descent and telescoping arguments, projection
geometry, projected-gradient mappings, residual certificates, strong
convexity, and linear convergence rates.

The main purpose of the entry is not only to formalize a single convergence
proof, but to separate the analytic, geometric, and algorithmic components
of first-order convergence arguments into reusable Isabelle/HOL layers.
The resulting library can be used as a basis for future formalizations of
constrained first-order optimization methods.
\end{abstract}

\tableofcontents

\section{Introduction}

First-order methods are among the basic algorithmic tools of convex
optimization.  Their convergence proofs are usually built from a small
number of reusable ingredients: first-order convexity inequalities, smooth
quadratic upper bounds, one-step descent estimates, telescoping arguments,
and, in the constrained case, projection inequalities.  This entry formalizes
these ingredients in Isabelle/HOL and combines them to obtain convergence
guarantees for gradient descent and projected-gradient descent.

The focus of the development is smooth convex optimization.  In the
unconstrained case, the main algorithmic object is the gradient step.  In
the constrained case, the main object is the projected-gradient step, together
with the associated projected-gradient mapping.  The projected-gradient
mapping plays the role of a constrained residual: its vanishing characterizes
a projected-gradient fixed point and, under convexity assumptions, a global
minimizer.  The entry also proves finite-horizon residual estimates, which
give an explicit approximate-stationarity certificate for projected-gradient
descent.

The development is intended as a reusable library rather than a standalone
formalization of one theorem.  The theories are organized so that the
algorithm-independent parts, such as descent and telescoping lemmas, are
separated from gradient-specific and projection-specific arguments.  This
separation is meant to make the entry useful for later formalizations of
other first-order methods.

The mathematical presentation follows standard references on convex
optimization and first-order methods, such as
\cite{Nesterov2004,Nesterov2018,Beck2017,Bertsekas1999}.  The entry also builds on the
existing Isabelle/HOL infrastructure for analysis, convexity, and projections.

\section{Main contributions}

The main contributions of the entry are as follows.

\begin{itemize}
  \item It provides reusable gradient and first-order convexity interfaces
  for smooth convex optimization in Isabelle/HOL.

  \item It separates analytic smoothness certificates from algorithmic
  convergence proofs through a quadratic smooth upper-bound interface and
  related line-descent formulations.

  \item It develops an algorithm-independent descent layer based on
  telescoping and averaging arguments, so that one-step progress estimates
  can be reused in convergence proofs for first-order methods.

  \item It formalizes convergence proofs for gradient descent and
  projected-gradient descent under smooth convex assumptions.

  \item It develops projection geometry and projected-gradient mapping
  residuals as reusable tools for constrained smooth convex optimization.

  \item It proves finite-horizon residual and epsilon-stationarity
  certificates for projected-gradient descent.

  \item It adds strong-convexity refinements, including distance-gap
  estimates and linear-rate convergence results.

  \item It provides concrete quadratic examples and a public theorem
  interface intended for downstream developments.
\end{itemize}

\section{Structure of the development}

The formalization is organized into several layers.

\paragraph{Analytic and convexity interfaces.}
The theories \texttt{Gradient\_Preliminaries} and
\texttt{Convex\_Differentiable} contain the basic gradient and first-order
convexity infrastructure used throughout the entry.  They isolate the
first-order inequalities needed in convergence proofs from the later
algorithmic arguments.

\paragraph{Smooth upper bounds and descent estimates.}
The theories \texttt{Smooth\_Convex} and \texttt{Lipschitz\_Smoothness}
develop the smoothness interface.  The central object is the standard
quadratic upper bound for smooth functions.  This upper bound is the main
tool used to derive one-step progress estimates for gradient descent and
projected-gradient descent.

\paragraph{Algorithm-independent descent tools.}
The theory \texttt{Abstract\_Descent} contains telescoping, summability,
and averaging lemmas.  These lemmas are deliberately separated from any
particular algorithm.  Once an algorithm supplies a suitable one-step
progress inequality, the same abstract descent layer can be used to obtain
finite-horizon convergence estimates.

\paragraph{Gradient descent.}
The theories \texttt{Gradient\_Descent}, \texttt{Gradient\_Descent\_Rates},
and \texttt{Gradient\_Descent\_Convergence} instantiate the smoothness and
descent infrastructure for the ordinary gradient method.  They prove the
standard sublinear function-value estimates for smooth convex minimization.

\paragraph{Projection geometry and projected-gradient descent.}
The theories \texttt{Projection\_Geometry} and
\texttt{Projection\_Optimization} collect the projection inequalities and
variational characterizations needed for constrained optimization.  The
theory \texttt{Projected\_Gradient\_Descent\_Convergence} then proves the
main convergence results for projected-gradient descent over a closed convex
set.

\paragraph{Projected-gradient mappings and residual certificates.}
The theories \texttt{Projected\_Gradient\_Mapping},
\texttt{Projected\_Gradient\_Mapping\_Rates}, and
\texttt{Projected\_Gradient\_Descent\_Residual\_Convergence} develop the
projected-gradient mapping as a constrained first-order residual.  This part
of the entry proves fixed-point and optimality characterizations, summability
and averaging estimates for the residual norm, and finite-time
epsilon-stationarity certificates.

\paragraph{Strong convexity and linear convergence.}
The theories \texttt{Strong\_Convex} and
\texttt{Projected\_Gradient\_Descent\_Linear\_Rate} formalize the
strong-convexity refinements.  These results upgrade the convex convergence
statements to linear-rate estimates under the usual strong-convexity and
smoothness assumptions.

\paragraph{Examples and public interface.}
The example theories instantiate the abstract interfaces on simple quadratic
objectives.  The theory \texttt{Main\_Results} collects the most important
theorem groups and is intended to be the main import target for downstream
developments.  The theory \texttt{Examples\_Using\_Main\_Results} demonstrates
how a client development can use the stable public theorem surface without
importing the lower-level proof files directly.  The final umbrella theory
\texttt{Projected\_Gradient\_Descent} imports the complete entry.

\section{Gradient and convexity interfaces}

The first layer of the entry provides the basic analytic vocabulary for
gradient-based optimization.  Instead of hard-coding a particular derivative
definition into each convergence proof, the development works with an
abstract gradient interface.  This makes the later results easier to reuse
with different concrete differentiability assumptions.

The convexity layer formalizes the first-order inequality for differentiable
convex functions.  Informally, for a convex differentiable function \(f\)
with gradient \(G\), one has
\[
  f(y) \geq f(x) + \langle G(x), y - x\rangle.
\]
This inequality is the main bridge between local first-order information and
global objective-value estimates.  It is used both in the unconstrained
gradient-descent proofs and in the constrained projected-gradient arguments.

\section{Smooth upper bounds and descent estimates}

Smoothness is represented through the usual quadratic upper bound.  In
informal notation, the smoothness interface states that
\[
  f(y) \leq f(x) + \langle G(x), y - x\rangle
        + \frac{L}{2}\|y-x\|^2.
\]
This inequality is the basic descent tool for first-order methods.  When
\(y\) is chosen as a gradient or projected-gradient step, the quadratic upper
bound gives a one-step decrease estimate.

The formalization keeps this smooth upper-bound interface separate from the
algorithmic convergence proofs.  This is useful because many first-order
methods share the same proof pattern: first prove a local one-step progress
inequality, then use an abstract telescoping argument to obtain a global
finite-horizon rate.

The entry also provides a Lipschitz-smoothness layer.  In particular, it
records the equivalence between the line-descent certificate and the smooth
upper-bound interface, and it includes a mean-value bridge showing how a
line mean-value certificate together with a Lipschitz gradient yields the
same smoothness interface with constant \(2L\).  This bridge is deliberately
separated from the algorithmic convergence layer, so that later refinements
can replace it by a sharper integral descent-lemma proof without changing
the projected-gradient convergence theorems.

\section{Gradient descent}

For unconstrained smooth convex minimization, the entry formalizes the
ordinary gradient step and proves the standard sublinear convergence
estimates.  The gradient-descent development serves two roles.  First, it
provides a basic formalization of a classical first-order method.  Second,
it acts as a simpler model for the later projected-gradient development:
the same smoothness and descent infrastructure is reused, but without the
additional projection geometry.

The convergence proof follows the standard structure.  The smooth upper bound
gives a one-step descent inequality.  Convexity relates the gradient term to
the objective gap.  Finally, a telescoping argument turns the one-step
inequality into a finite-horizon convergence rate.

\section{Projection geometry and projected-gradient descent}

The constrained part of the entry is built around metric projection onto a
closed convex set.  The projection step is characterized by a variational
inequality.  Informally, if \(p\) is the projection of \(z\) onto a closed
convex set \(C\), then
\[
  \langle z - p, u - p\rangle \leq 0
\]
for all \(u \in C\).  This inequality is the geometric replacement for the
unconstrained identity used in gradient descent.

Projected-gradient descent combines the smooth upper bound with this
projection inequality.  The projected step has the form
\[
  x^+ = \operatorname{proj}_C(x - \alpha G(x)).
\]
The formal proof separates the projection geometry from the descent
argument.  This separation is important for reusability: the same projection
lemmas can be used independently of the particular convergence theorem in
which they appear.

The entry proves function-value convergence guarantees for projected-gradient
descent under the standard smooth convex assumptions and a suitable
stepsize condition.

\section{Projected-gradient mappings}

The projected-gradient mapping is the central constrained residual used in
this entry.  For stepsize \(\alpha > 0\), it is informally given by
\[
  \mathcal{G}_{\alpha}(x)
    = \frac{1}{\alpha}
      \left(x - \operatorname{proj}_C(x - \alpha G(x))\right).
\]
In the unconstrained case, this residual reduces to the ordinary gradient.
In the constrained case, it measures the failure of \(x\) to be a fixed point
of the projected-gradient step.

The entry formalizes the connection between zero projected-gradient mapping,
projected-gradient fixed points, variational inequalities, and global
optimality for convex problems.  It also proves finite-horizon residual
estimates.  A typical consequence is that, among the first \(N\) iterates
of projected-gradient descent, at least one iterate has small
projected-gradient residual, with an explicit bound depending on the initial
objective gap, the stepsize, and \(N\).

This residual layer is one of the main reusable components of the entry.
It packages convergence not only as a function-value guarantee, but also as
an approximate first-order stationarity certificate for constrained smooth
convex optimization.

\section{Strong convexity and linear convergence}

The entry also formalizes a strong-convexity layer.  Strong convexity gives
quantitative control of the distance to the minimizer in terms of objective
gaps and first-order certificates.  Under the usual combination of smoothness
and strong convexity assumptions, projected-gradient descent satisfies a
linear convergence estimate.

This part of the development complements the sublinear convex theory.  The
convex results establish the basic \(O(1/N)\)-type finite-horizon behavior,
while the strongly convex results provide the corresponding linear-rate
refinement.

\section{Examples}

The example theories instantiate the abstract interfaces on quadratic
objectives.  These examples serve as compact regression tests for the
library: they show how the smoothness, convexity, projection, residual, and
linear-rate interfaces can be used in concrete problems.

The constrained quadratic examples include both a nonnegative half-line
constraint and a bounded interval constraint.  The nonnegative half-line
example gives a simple setting in which projection geometry,
projected-gradient mappings, function-value estimates, and
epsilon-stationarity certificates can all be instantiated explicitly.

The bounded interval example further illustrates the reusable projection
layer: the convergence and residual results are obtained through the general
closed-convex projection interface, rather than through an ad hoc
closed-form projection calculation.

\section{Related work}

The Isabelle/HOL libraries already provide substantial infrastructure for
analysis, convex sets, convex functions, and metric projections.  This entry
builds on that background and develops an algorithmic layer for first-order
methods in smooth convex optimization.

The AFP entry \emph{Unconstrained Optimization} formalizes minimizers,
strict and isolated local minimizers, and first- and second-order optimality
conditions for unconstrained problems \cite{UnconstrainedOptimizationAFP}.
The present development is complementary: it focuses not on local optimality
conditions for unconstrained problems, but on convergence proofs for gradient
descent and projected-gradient descent, together with projection geometry,
projected-gradient mappings, and residual certificates for constrained smooth
convex optimization.

Outside Isabelle/HOL, Li et al. formalize convergence rates of several
first-order algorithms for convex optimization in Lean4
\cite{FirstOrderMethodsLean}.  Their work develops Lean infrastructure for
gradients, subgradients, differentiable convex functions, and convergence
proofs for a range of first-order algorithms.  The present entry contributes
complementary Isabelle/HOL/AFP infrastructure, with particular emphasis on
projection geometry, projected-gradient descent, projected-gradient mappings,
and residual certificates for constrained smooth convex optimization.

\section{Conclusion}

This entry formalizes a reusable Isabelle/HOL layer for smooth convex
first-order optimization.  Its main contributions are the separation of
gradient, smoothness, descent, projection, and residual arguments; convergence
proofs for gradient descent and projected-gradient descent; projected-gradient
mapping optimality and residual estimates; and strong-convexity linear-rate
refinements.

The structure of the development is intended to support future extensions.
Natural next steps include proximal-gradient methods, conditional-gradient
methods, accelerated methods, additional constrained examples such as box or
polyhedral constraints, and sharper analytic bridges from Lipschitz-gradient assumptions to the quadratic
upper-bound interface, for example via integral descent-lemma arguments along
line segments.

\bibliographystyle{plain}
\bibliography{root}

\end{document}